Individual and Differential Harm in Redistricting

### 2025-08-25 paper/harm.qmd
remove use of \indep
cuts

---
title: Individual and Differential Harm in Redistricting
date: last-modified
authors:
  - name: Cory McCartan
    affiliations:
      - name: Pennsylvania State University
        department: Department of Statistics
        address: 461 Pollock Road
        city: University Park
        region: PA
        postal-code: 16801
    email: mccartan@psu.edu
    url: https://corymccartan.com/
    corresponding: true
  - name: Christopher T. Kenny
    affiliations:
    - name: Harvard University
      department: Department of Government
      address: 1737 Cambridge Street
      city: Cambridge
      region: MA
      postal-code: "02138 "
    email: christopherkenny@fas.harvard.edu
    url: https://christophertkenny.com
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups,
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating
  the impacts of a districting plan on individual voters and the groups to which they belong.
  We consider a voter harmed if their chosen candidate is not elected under the current plan,
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework
  through analyses of partisan gerrymandering in New Jersey, voting rights
  litigation in Alabama, and racial dynamics of Boston City Council elections.
thanks: |
  The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King,
  Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and
  participants of the Fall 2021 Harvard American Politics Research Workshop,
  Fall 2022 Harvard Applied Statistics Workshop,
  and Imai Research Group for helpful comments and discussion.
  Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
format:
    cmc-article-pdf:
        journal:
            blinded: true
        fontsize: 11pt
        biblio-style: apsr
        include-in-header: "header.tex"
editor_options:
  markdown:
    wrap: sentence
---

@@

The academic study of gerrymandering has not reflected this diversity.
Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
This literature focuses on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

@@

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
The harm measure also captures or generalizes the three *Gingles* preconditions for vote dilution litigation under the Voting Rights Act of 1965, as discussed further in [Appendix @sec-connections].

@@

Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$, as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.

In general, $q$ harms some individuals versus $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals, versus $q$.
A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics.

To allow for such averaging, we replace the fixed election $v$, counterfactual districting plan $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.

>>>>

---
title: Individual and Differential Harm in Redistricting
date: last-modified
authors:
  - name: Cory McCartan
    affiliations:
      - name: Pennsylvania State University
        department: Department of Statistics
        address: 461 Pollock Road
        city: University Park
        region: PA
        postal-code: 16801
    email: mccartan@psu.edu
    url: https://corymccartan.com/
    corresponding: true
  - name: Christopher T. Kenny
    affiliations:
    - name: Harvard University
      department: Department of Government
      address: 1737 Cambridge Street
      city: Cambridge
      region: MA
      postal-code: "02138 "
    email: christopherkenny@fas.harvard.edu
    url: https://christophertkenny.com
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups,
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating
  the impacts of a districting plan on individual voters and the groups to which they belong.
  We consider a voter harmed if their chosen candidate is not elected under the current plan,
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework
  through analyses of partisan gerrymandering in New Jersey, voting rights
  litigation in Alabama, and racial dynamics of Boston City Council elections.
thanks: |
  The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King,
  Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and
  participants of the Fall 2021 Harvard American Politics Research Workshop,
  Fall 2022 Harvard Applied Statistics Workshop,
  and Imai Research Group for helpful comments and discussion.
  Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
format:
    cmc-article-pdf:
        journal:
            blinded: true
        keep-tex: true
        fontsize: 11pt
        biblio-style: apsr
        include-in-header: "header.tex"
editor_options:
  markdown:
    wrap: sentence
---

@@

The academic study of gerrymandering has not reflected this diversity.
Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @thomas2013estimating; @katz2020].
A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
This literature focuses on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

@@

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [e.g., @1962baker; @issakarl98].
This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [e.g., @2004vieth].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
The harm measure also captures or generalizes the three *Gingles* preconditions for vote dilution litigation under the Voting Rights Act of 1965, as discussed further in [Appendix @sec-connections].

@@

Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$, since $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
In general, $q$ harms some individuals versus $\tilde q$, and $\tilde q$ harms a distinct set of individuals versus $q$.
A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
Researchers are generally interested in outcomes over multiple elections and a much larger set of counterfactual plans.
<!-- Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics. -->

To allow for this broader scope, we replace the fixed election $v$, counterfactual districting plan $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
The random variable $V$ represents a statistical model for possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
The random variable $\tilde Q$ represents a model for potential counterfactual districting plans.
Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
